---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
    toc: yes
---

## Introduction

It is now possible to collect a large amount of data about personal movement
using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone
Up. These type of devices are part of the “quantified self” movement – a group
of enthusiasts who take measurements about themselves regularly to improve
their health, to find patterns in their behavior, or because they are tech geeks.
But these data remain under-utilized both because the raw data are hard to
obtain and there is a lack of statistical methods and software for processing and
interpreting the data.

This assessment makes use of data from a personal activity monitoring device.
This device collects data at 5 minute intervals through out the day. The data
consists of two months of data from an anonymous individual collected during
the months of October and November, 2012 and include the number of steps
taken in 5 minute intervals each day.

## Data

The data can be downloaded from the course web site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a 
total of 17,568 observations in this dataset.

For convience the raw dataset is included in this repository.

## Loading and preprocessing the data

Ensure that we show all our working:
```{r setoptions, echo=TRUE}
require(knitr, quietly = TRUE)
opts_chunk$set(echo = TRUE, cache = TRUE)
```

Load data into a data frame:
```{r loaddata}
require(utils, quietly = TRUE)

## unzip overwriting existing directory to ensure clean setup
if (!file.exists("activity.csv")) {
    # download archive into local directory
    zipFileName <- file.path("activity.zip")
    if (!file.exists(zipFileName)) {
        zipUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(zipUrl, destfile = zipFileName, method = "curl", mode = "wb")
        print(paste(Sys.time(), "archive downloaded"))
    }
    unzip(zipFileName, overwrite = FALSE)
    print(paste(Sys.time(), "archive unpacked"))
}

## load into data frame and convert date column to date
data <- read.csv("activity.csv", stringsAsFactors = FALSE)
# make new column datetime using date and 5 minute intervals
data <- transform(data, 
                  datetime = strptime(
                                paste(date, formatC(interval, flag = "0", width = 4)), 
                                format = "%Y-%m-%d %H%M"), 
                  date = as.Date(data$date, "%Y-%m-%d")
                  )
```

## What is mean total number of steps taken per day?

The following histogram shows the total number of steps taken each day during 
the two month period from `r format(min(data$date), "%a %b %d, %Y")`
to `r format(max(data$date), "%a %b %d, %Y")`. This ignores days for which no 
data was recorded.

Aggregate the total number of steps per day:
```{r aggregate}
dailyTotals <- aggregate(steps ~ date, data, FUN = sum)
```

Plot the total number of steps per day as a histogram:
```{r histogram}
require(magrittr, quietly = TRUE)
require(ggplot2, quietly = TRUE)
require(scales, quietly = TRUE)
dailyTotals %>%
    ggplot(aes(steps)) + 
    geom_histogram(binwidth = 1000, fill = "purple", colour = "black", alpha = 0.7) +
    theme_light(base_family = "Avenir", base_size = 11) +
    scale_y_continuous(breaks = pretty_breaks(10)) +
    labs(x = "Steps") +
    labs(y = "Frequency") +
    ggtitle("Histogram of Total Steps per Day")
```

The mean number of steps was `r prettyNum(round(mean(dailyTotals$steps)), big.mark=",")` 
per day. _(Rounded up from `r prettyNum(mean(dailyTotals$steps), big.mark=",")` as fractional 
steps do not make sense.)_  
The median number of steps was `r prettyNum(median(dailyTotals$steps), big.mark=",")` per day. 

## What is the average daily activity pattern?

The following is a time series (line) plot of the 5-minute intervals 
and the average number of steps taken, averaged across all days.

```{r timeseries}
intervalTotals <- aggregate(steps ~ interval, data, FUN = mean)

require(ggplot2, quietly = TRUE)
require(scales, quietly = TRUE)
intervalTotals %>%
    ggplot(aes(interval, steps)) + 
    # geom_point(size = 3) +
    geom_line(color = "purple") +
    theme_light(base_family = "Avenir", base_size = 11) +
    scale_x_discrete(breaks = pretty_breaks(20)) +
    scale_y_continuous(breaks = pretty_breaks(10)) +
    labs(x = "Interval") +
    labs(y = "Steps") +
    theme(legend.position = "none") +
    ggtitle("5-minute Interval Time Series")
```

The 5-minute interval, which on average across all the days in the dataset, contains the maximum number of steps is `r subset(intervalTotals, subset = steps == max(steps), select = "interval")`.

## Imputing missing values

Note that there are a number of days/intervals where there are missing
values (coded as `NA`). The presence of missing days may introduce
bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with `NA`s)

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

4. Make a histogram of the total number of steps taken each day and Calculate and report the **mean** and **median** total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

## Are there differences in activity patterns between weekdays and weekends?


For this part the `weekdays()` function may be of some help here. Use
the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

1. Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).